---
title: "Marked custom-Stan sampling for an ordinal Markov random field"
format: html
vignette: >
  %\VignetteIndexEntry{Marked custom-Stan sampling for an ordinal Markov random field}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
---

```{r}
library(PDMPSamplersR)
```

The OMRF pseudolikelihood is a sum over persons. For person (n), each node is
a categorical-logit conditional whose non-reference logits are affine in the
threshold and interaction parameters. The marked observation is therefore one
person, including all node conditionals for that person.

For node (j), write the affine logit map as (B_{nj}\theta). The
categorical-logit Hessian has operator norm at most (1/2), so

\[
w_n = \frac12 \sum_j \lVert B_{nj}\rVert_{op}^2
\]

is a global nonnegative curvature coefficient. `omrf_residual_envelope()`
computes these coefficients once from the ordinal scores and parameter-block
names. Julia combines them with flow-specific displacement and velocity bounds
inside its typed `SeparableResidualEnvelope`; no R callback runs in the event
loop. Finite diagnostics below exercise the implementation, while the global
categorical-logit curvature inequality is the guarantee.

## Deterministic example and subset hook

```{r}
N <- 12L
P <- 3L
K <- 3L
X <- matrix(
  c(0, 1, 2,  2, 0, 1,  1, 2, 0,  0, 2, 1,
    1, 0, 2,  2, 1, 0,  0, 1, 1,  1, 2, 2,
    2, 0, 0,  0, 2, 2,  1, 1, 0,  2, 2, 1),
  nrow = N, byrow = TRUE
)
stan_data <- list(
  N = N, P = P, K = K, X = X, seen = rep(K, P), prior_only = 0L,
  prior_interaction_sd = 0.5,
  prior_threshold_alpha = 2,
  prior_threshold_beta = 2
)
prior_data <- stan_data
prior_data$prior_only <- 1L

stan_file <- system.file(
  "stan", "omrf", "omrf_marked.stan", package = "PDMPSamplersR"
)
pdmp_subsample_hpp_path()
```

The Stan source declares `pdmp_get_subsample_size()` and
`pdmp_get_subsample_index()`. `compile_pdmp_stan_model()` installs the packaged
header and the required compiler flags. Its thread-local buffer selects three
execution modes from one persistent library:

- an empty subset plus `prior_only = 0` evaluates the full likelihood;
- `prior_only = 1` evaluates only the ordinary prior;
- a nonempty subset evaluates the unscaled sum of those persons.

Julia applies `N / m` exactly once. The accepted selected gradient is retained
for reflection, and every later proposal draws a fresh subset.

```{r}
envelope <- omrf_residual_envelope(
  X = stan_data$X,
  seen = stan_data$seen,
  thresholds = "thresholds_0",
  interactions = "interactions_0"
)
anchor <- rep(0, 9L)
marked <- stan_marked_subsampling(
  n_observations = stan_data$N,
  subsample_size = 3L,
  prior_standata = prior_data,
  residual_envelope = envelope,
  anchor = anchor
)
```

The same generic constructor works for another custom Stan likelihood only
when it receives a certified residual envelope. The subset hook alone is an
efficient selected-gradient oracle, not a proof of envelope domination.

## Full and marked fits

The following managed check is disabled during routine vignette builds because
it compiles Stan and runs replicated posterior chains. Set
`PDMPSAMPLERSR_RUN_OMRF_VIGNETTE=true` to execute it. The benchmark driver
`benchmarks/omrf_marked.R` contains the longer (N=200,500),
(P=10,20,30) timing and ESS/second study.

```{r}
run_checks <- identical(
  tolower(Sys.getenv("PDMPSAMPLERSR_RUN_OMRF_VIGNETTE")), "true"
)
```

```{r}
#| eval: !expr run_checks
compiled <- compile_pdmp_stan_model(stan_file)
mapping <- stan_parameter_mapping(compiled, stan_data)
stopifnot(identical(mapping$name[1:6], paste0("thresholds_0.", 1:6)))

edges <- rbind(c(1L, 2L), c(1L, 3L), c(2L, 3L))
person_loglik <- function(q, x) {
  thresholds <- matrix(q[1:6], nrow = 2L)
  interactions <- matrix(0, P, P)
  for (e in seq_len(nrow(edges))) {
    interactions[edges[e, 1L], edges[e, 2L]] <- q[6L + e]
    interactions[edges[e, 2L], edges[e, 1L]] <- q[6L + e]
  }
  result <- 0
  for (j in seq_len(P)) {
    if (x[j] > 0L) result <- result + thresholds[x[j], j]
    field <- sum(x * interactions[j, ])
    eta <- c(0, thresholds[, j] + (1:2) * field)
    result <- result - (max(eta) + log(sum(exp(eta - max(eta)))))
  }
  result + sum(vapply(seq_len(nrow(edges)), function(e) {
    2 * interactions[edges[e, 1L], edges[e, 2L]] *
      x[edges[e, 1L]] * x[edges[e, 2L]]
  }, numeric(1)))
}
full_loglik <- function(q) {
  sum(vapply(seq_len(N), function(n) person_loglik(q, X[n, ]), numeric(1)))
}
q <- seq(-0.2, 0.25, length.out = 9L)
stopifnot(isTRUE(all.equal(
  full_loglik(q),
  sum(vapply(seq_len(N), function(n) person_loglik(q, X[n, ]), numeric(1)))
)))

subset <- c(2L, 5L, 9L)
diagnostic <- stan_marked_diagnostics(
  compiled, stan_data, marked, position = q, subset = subset,
  velocity = seq(0.4, 1.2, length.out = 9L), flow = "ZigZag"
)
step <- 1e-6
explicit_selected <- vapply(seq_along(q), function(j) {
  plus <- minus <- q
  plus[j] <- plus[j] + step
  minus[j] <- minus[j] - step
  -(sum(vapply(subset, function(n) person_loglik(plus, X[n, ]), numeric(1))) -
      sum(vapply(subset, function(n) person_loglik(minus, X[n, ]), numeric(1)))) /
    (2 * step)
}, numeric(1))
stopifnot(
  isTRUE(all.equal(diagnostic$selected_likelihood_gradient,
                   explicit_selected, tolerance = 2e-7)),
  diagnostic$selected_scale == N / length(subset),
  isTRUE(all.equal(
    diagnostic$marked_gradient,
    diagnostic$deterministic_gradient +
      diagnostic$selected_scale * diagnostic$residual_gradient,
    tolerance = 1e-12)),
  max(abs(diagnostic$residual_gradient)) > 1e-4
)
for (flow in c("ZigZag", "BouncyParticle", "AdaptiveBoomerang")) {
  flow_check <- stan_marked_diagnostics(
    compiled, stan_data, marked, position = q, subset = subset,
    velocity = diagnostic$residual_gradient, flow = flow)
  stopifnot(
    flow_check$residual_rate > 0,
    flow_check$residual_rate <=
      flow_check$envelope_rate * (1 + 1e-10) + 1e-10
  )
}

anchor_check <- stan_marked_diagnostics(
  compiled, stan_data, marked, position = anchor, subset = subset,
  velocity = rep(1, 9L), flow = "BouncyParticle"
)
stopifnot(isTRUE(all.equal(
  anchor_check$marked_gradient, anchor_check$full_gradient,
  tolerance = 1e-10
)))

slab <- independent_slab_density(
  1 / (sqrt(2 * pi) * stan_data$prior_interaction_sd),
  coef = "interactions_0"
)
common <- list(
  path_to_stanmodel = compiled, standata = stan_data,
  flow = "ZigZag", algorithm = "GridThinningStrategy",
  T = 300, grid_n = 15L, grid_t_max = 0.5,
  sticky = TRUE, can_stick = "interactions_0",
  model_prior = betabernoulli(1, 4), slab_prior = slab,
  show_progress = FALSE, materialize = FALSE
)
seeds <- 901:903
full_fits <- lapply(seeds, function(seed) {
  do.call(pdmp_sample_from_stanmodel, c(common, list(seed = seed)))
})
marked_fits <- lapply(seeds, function(seed) {
  do.call(pdmp_sample_from_stanmodel,
          c(common, list(marked_subsampling = marked, seed = seed)))
})
replicated_summary <- function(fits) {
  run_means <- do.call(cbind, lapply(fits, mean))
  run_variances <- do.call(cbind, lapply(fits, function(fit) diag(var(fit))))
  run_ess <- do.call(cbind, lapply(fits, ess))
  runs <- length(fits)
  within_mcse2 <- rowSums(run_variances / run_ess) / runs^2
  between_mcse2 <- apply(run_means, 1L, stats::var) / runs
  list(
    mean = rowMeans(run_means),
    mcse = sqrt(pmax(within_mcse2, between_mcse2)),
    minimum_ess = min(run_ess)
  )
}
full_summary <- replicated_summary(full_fits)
marked_summary <- replicated_summary(marked_fits)
standardized_difference <- abs(full_summary$mean - marked_summary$mean) /
  sqrt(full_summary$mcse^2 + marked_summary$mcse^2)
stopifnot(
  full_summary$minimum_ess > 20,
  marked_summary$minimum_ess > 20,
  all(is.finite(standardized_difference)),
  max(standardized_difference) < 4.5
)
for (marked_fit in marked_fits) {
  counters <- attr(marked_fit, "marked_context_counters")[[1L]]
  if (is.environment(counters)) counters <- as.list(counters)
  stopifnot(
    counters$persons_evaluated ==
      marked$subsample_size * counters$selected_gradient_calls,
    counters$sampling_full_gradient_calls == 0L,
    counters$sampling_model_constructions == 0L,
    counters$sampling_data_constructions == 0L
  )
}
```

`selected_gradient_calls` counts both the current-position and anchor
evaluations needed by each residual call, so `persons_evaluated` is exactly
`m * selected_gradient_calls`. `initialization_full_gradient_calls` records the
anchor preparation; `sampling_full_gradient_calls` must remain zero. Likewise,
the initialization model/data construction counts are fixed before sampling,
and both sampling construction deltas must remain zero.
