---
title: "Sampling from Stan models with PDMPSamplersR"
format: html
vignette: >
  %\VignetteIndexEntry{Sampling from Stan models with PDMPSamplersR}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(PDMPSamplersR)
library(tibble)
library(ggplot2)
library(patchwork)
library(fs)

julia_available <- tryCatch(
    { JuliaCall::julia_setup(verbose = FALSE); TRUE },
    error = function(e) FALSE
)
knitr::opts_chunk$set(eval = julia_available)

theme_set(
  theme_bw(base_size = 16) +
    theme(
      geom = element_geom(
        pointshape = 21, pointsize = 5,
        fill = alpha("grey", .5)
      )
    )
)

stan_model_dir <- system.file("stan", "models", package = "PDMPSamplersR")
if (!nzchar(stan_model_dir)) {
  stan_model_dir <- file.path("..", "inst", "stan", "models")
}
```

```{r}
#| label: setup_shown
#| include: true
#| eval: false
library(PDMPSamplersR)
library(tibble)
library(ggplot2)
library(patchwork)
library(fs)

# some personal recommendations for nicer-looking plots
# requires ggplot2 >= 3.4.0 for element_geom()
theme_set(
  theme_bw(base_size = 16) +
    theme(
      geom = element_geom(pointshape = 21, pointsize = 5, fill = alpha("grey", .5))
    )
)

stan_model_dir <- system.file("stan", "models", package = "PDMPSamplersR")
```


This vignette demonstrates how to use `PDMPSamplersR` to sample from
Stan models using piecewise deterministic Markov processes (PDMPs). We
cover:

1. Multivariate normal targets with the ZigZag sampler
2. Grid-based thinning strategies
3. Variable selection via spike-and-slab priors (Bernoulli and
   beta-binomial)
4. Logistic regression
5. A Bayesian t-test with Bayes factor computation

## Multivariate normal with ZigZag

We start by sampling from a 5-dimensional multivariate normal
distribution. The mean and covariance are drawn at random and passed
both as the Stan data and as the flow parameters.

```{r}
#| label: mvnormal-zigzag
set.seed(123)
d <- 5
mean_vec   <- rnorm(d, 0, 2)
cov_matrix <- stats::rWishart(1, df = d + 2, Sigma = diag(d))[, , 1]
stan_data  <- list(N = d, mu = mean_vec, sigma = cov_matrix)

data_path <- tempfile(fileext = ".json")
write_stan_json(stan_data, data_path)

model_path <- path(stan_model_dir, "mvnormal.stan")

result <- pdmp_sample_from_stanmodel(
  model_path, data_path,
  flow = "ZigZag", T = 10000,
  flow_mean = mean_vec, flow_cov = cov_matrix
)
```

Compare estimated and true means and covariances:

```{r}
#| label: mvnormal-zigzag-plot
#| fig-width: 8
#| fig-height: 5
est_mean <- mean(result)
est_cov  <- cov(result)

tib <- tibble(
  truth    = c(mean_vec, cov_matrix[lower.tri(cov_matrix, diag = TRUE)]),
  estimate = c(est_mean, est_cov[lower.tri(est_cov, diag = TRUE)]),
  what     = rep(c("mu", "sigma"), times = c(d, d * (d + 1) / 2))
)

ggplot(tib, aes(x = truth, y = estimate)) +
  geom_abline(slope = 1, intercept = 0, color = "grey", alpha = .8) +
  geom_point() +
  facet_wrap(~what, scales = "free") +
  labs(title = "ZigZag PDMP", x = "True value", y = "Estimated value")
```

## Multivariate normal with grid thinning

The same target, but using the `GridThinningStrategy` algorithm.

```{r}
#| label: mvnormal-grid
result2 <- pdmp_sample_from_stanmodel(
  model_path, data_path,
  flow = "ZigZag", algorithm = "GridThinningStrategy", T = 10000
)
```

```{r}
#| label: mvnormal-grid-plot
#| fig-width: 8
#| fig-height: 5
est_mean <- mean(result2)
est_cov  <- cov(result2)

tib <- tibble(
  truth    = c(mean_vec, cov_matrix[lower.tri(cov_matrix, diag = TRUE)]),
  estimate = c(est_mean, est_cov[lower.tri(est_cov, diag = TRUE)]),
  what     = rep(c("mu", "sigma"), times = c(d, d * (d + 1) / 2))
)

ggplot(tib, aes(x = truth, y = estimate)) +
  geom_abline(slope = 1, intercept = 0, color = "grey", alpha = .8) +
  geom_point() +
  facet_wrap(~what, scales = "free") +
  labs(title = "ZigZag PDMP (Grid Thinning)", x = "True value", y = "Estimated value")
```

## Spike-and-slab variable selection (Bernoulli prior)

When the covariance matrix is diagonal, variable selection with sticky
ZigZag is exact. We set different prior inclusion probabilities for each
dimension and check recovery.

```{r}
#| label: spike-slab-bernoulli
set.seed(123)
d <- 12
mean_vec   <- rep(0, d)
cov_matrix <- stats::rWishart(1, df = d + 2, Sigma = diag(d))[, , 1]
cov_matrix <- diag(diag(cov_matrix))
stan_data  <- list(N = d, mu = mean_vec, sigma = cov_matrix)

data_path <- tempfile(fileext = ".json")
write_stan_json(stan_data, data_path)

prior_incl_prob <- seq(0.1, 0.9, length.out = d)

result3 <- pdmp_sample_from_stanmodel(
  model_path, data_path,
  flow = "ZigZag", T = 10000,
  algorithm = "GridThinningStrategy",
  sticky = TRUE, can_stick = rep(TRUE, d),
  model_prior = bernoulli(prior_incl_prob),
  parameter_prior = dnorm(rep(0, d), mean_vec, sqrt(diag(cov_matrix)))
)
```

```{r}
#| label: spike-slab-bernoulli-plot
#| fig-width: 6
#| fig-height: 5
tib <- tibble(
  true = prior_incl_prob,
  est  = inclusion_probs(result3)
)

ggplot(tib, aes(x = true, y = est)) +
  geom_abline(slope = 1, intercept = 0, color = "grey", alpha = .8) +
  geom_point() +
  labs(
    title = "Variable selection (Bernoulli prior)",
    x = "True inclusion probability",
    y = "Estimated inclusion probability"
  )
```

### Non-diagonal covariance (biased case)

When the covariance matrix is not diagonal, the spike-and-slab approach
produces biased results because the conditional distributions change
when variables are excluded. This is a known limitation.

```{r}
#| label: spike-slab-biased
set.seed(123)
d <- 12
mean_vec   <- rep(0, d)
cov_matrix <- stats::rWishart(1, df = d + 2, Sigma = diag(d))[, , 1]
stan_data  <- list(N = d, mu = mean_vec, sigma = cov_matrix)

data_path <- tempfile(fileext = ".json")
write_stan_json(stan_data, data_path)

prior_incl_prob <- seq(0.1, 0.9, length.out = d)

result4 <- pdmp_sample_from_stanmodel(
  model_path, data_path,
  flow = "ZigZag", T = 10000,
  algorithm = "GridThinningStrategy",
  sticky = TRUE, can_stick = rep(TRUE, d),
  model_prior = bernoulli(prior_incl_prob),
  parameter_prior = dnorm(rep(0, d), mean_vec, sqrt(diag(cov_matrix)))
)
```

```{r}
#| label: spike-slab-biased-plot
#| fig-width: 6
#| fig-height: 5
tib <- tibble(
  true = prior_incl_prob,
  est  = inclusion_probs(result4)
)

ggplot(tib, aes(x = true, y = est)) +
  geom_abline(slope = 1, intercept = 0, color = "grey", alpha = .8) +
  geom_point() +
  labs(
    title = "Variable selection (non-diagonal — biased)",
    x = "True inclusion probability",
    y = "Estimated inclusion probability"
  )
```

## Spike-and-slab with beta-binomial prior

Instead of independent Bernoulli priors on inclusion, we can use a
beta-binomial prior which induces a common inclusion probability across
all dimensions.

```{r}
#| label: spike-slab-betabinom
set.seed(123)
d <- 7
mean_vec   <- rep(0, d)
cov_matrix <- stats::rWishart(1, df = d + 2, Sigma = diag(d))[, , 1]
cov_matrix <- diag(diag(cov_matrix))
stan_data  <- list(N = d, mu = mean_vec, sigma = cov_matrix)

data_path <- tempfile(fileext = ".json")
write_stan_json(stan_data, data_path)

mp <- betabernoulli(1, 1)
prior_incl_prob <- rep(mp$a / (mp$a + mp$b), d)

result5 <- pdmp_sample_from_stanmodel(
  model_path, data_path,
  flow = "ZigZag", T = 10000,
  algorithm = "GridThinningStrategy",
  sticky = TRUE, can_stick = rep(TRUE, d),
  model_prior = mp,
  parameter_prior = dnorm(rep(0, d), mean_vec, sqrt(diag(cov_matrix)))
)
```

```{r}
#| label: spike-slab-betabinom-helpers
beta_binomial_pmf <- function(k, d, a, b, log = FALSE,
                              individual_model = FALSE) {
  log_binom  <- lfactorial(d) - lfactorial(k) - lfactorial(d - k)
  log_beta_n <- lbeta(a + k, b + d - k)
  log_beta_d <- lbeta(a, b)
  log_prob   <- (!individual_model) * log_binom + log_beta_n - log_beta_d
  if (log) log_prob else exp(log_prob)
}
```

```{r}
#| label: spike-slab-betabinom-plot
#| fig-width: 10
#| fig-height: 5
samples <- discretize(result5)

tib_incl <- tibble(
  true = prior_incl_prob,
  est  = inclusion_probs(result5)
)
plt_incl <- ggplot(tib_incl, aes(x = true, y = est)) +
  geom_abline(slope = 1, intercept = 0, color = "grey", alpha = .8) +
  geom_point() +
  labs(
    title = "Inclusion probabilities",
    x = "True inclusion probability",
    y = "Estimated inclusion probability"
  )

model_indicators <- apply(1L * (samples != 0), 1L, paste0, collapse = "")
model_freq  <- table(model_indicators)
model_probs <- model_freq / sum(model_freq)
model_sizes <- nchar(gsub("0", "", names(model_freq), fixed = TRUE))

true_log_probs <- beta_binomial_pmf(
  model_sizes, d, mp$a, mp$b,
  log = TRUE, individual_model = TRUE
)
est_log_probs <- log(unname(c(model_probs)))

tib_model <- tibble(
  true_log = true_log_probs,
  est_log  = est_log_probs
)
plt_model <- ggplot(tib_model, aes(x = true_log, y = est_log)) +
  geom_abline(slope = 1, intercept = 0, color = "grey", alpha = .8) +
  geom_point() +
  labs(
    title = "Model probabilities",
    x = "True log model probability",
    y = "Estimated log model probability"
  )

plt_incl + plt_model
```

## Logistic regression with variable selection

We simulate binary data from a logistic model with 10 predictors, only
2 of which are active. We compare the PDMP posterior to a frequentist
GLM fit.

```{r}
#| label: logistic-regression
set.seed(123)
n <- 200
d <- 10
d_on <- floor(0.2 * d)
idx_on <- sample(1:d, d_on, FALSE)
intercept <- -3
beta <- rep(0, d)
beta[idx_on] <- rnorm(d_on, 0, 2)
X <- scale(matrix(rnorm(n * d), n, d))
linpred <- intercept + X %*% beta
y <- c(runif(n) <= plogis(linpred))
fit <- stats::glm(y ~ X, family = binomial())

model_path_lr <- path(stan_model_dir, "logistic_regression.stan")

sd_prior  <- 10
stan_data <- list(N = n, D = d, X = X, y = y, sd_prior = sd_prior)
data_path <- tempfile(fileext = ".json")
write_stan_json(stan_data, data_path)

result_lr <- pdmp_sample_from_stanmodel(
  model_path_lr, data_path,
  flow = "ZigZag", T = 20000,
  algorithm = "GridThinningStrategy",
  sticky = TRUE,
  can_stick = c(FALSE, rep(TRUE, d)),
  model_prior = bernoulli(0.5),
  parameter_prior = dnorm(rep(0, d + 1), 0, sd_prior)
)
```

```{r}
#| label: logistic-regression-plot
#| fig-width: 8
#| fig-height: 8
samples <- discretize(result_lr)

tib_lr <- tibble(
  true   = rep(c(intercept, beta), 2),
  est    = c(mean(result_lr), coef(fit)),
  how    = rep(c("Bayes", "Frequentist"), each = d + 1),
  p_incl = c(inclusion_probs(result_lr), rep(1, d + 1)),
  param  = rep(c("intercept", rep("beta", d)), 2)
)

plt_recovery <- ggplot(tib_lr, aes(x = true, y = est, fill = how,
                                   shape = param)) +
  geom_abline(slope = 1, intercept = 0, color = "grey", alpha = .8) +
  geom_point(alpha = .5) +
  scale_shape_manual(values = c(21, 24)) +
  facet_wrap(~how) +
  labs(title = "Parameter recovery", x = "True value", y = "Estimate")

plt_incl <- subset(tib_lr, how == "Bayes" & param == "beta") |>
  ggplot(aes(x = est, y = p_incl)) +
  geom_point() +
  labs(
    title = "Inclusion probability vs effect size",
    x = "Posterior mean", y = "Inclusion probability"
  )

plt_recovery / plt_incl
```

## Bayesian t-test with Bayes factor

Finally, we perform a Bayesian two-sample t-test and compare the PDMP
Bayes factor to the analytic result from the `BayesFactor` package.

```{r}
#| label: ttest
set.seed(123)
nx <- 30
ny <- 35
mux <- 0
muy <- 0.5
sigma2 <- 2.1

x1 <- rnorm(nx, mux, sqrt(sigma2))
x2 <- rnorm(ny, muy, sqrt(sigma2))

rscale <- 1.0
analytic_results <- BayesFactor::ttestBF(x = x1, y = x2, rscale = rscale)

model_path_tt <- path(stan_model_dir, "ttest.stan")
stan_data <- list(nx = nx, ny = ny, x = x1, y = x2, rscale = rscale)
data_path <- tempfile(fileext = ".json")
write_stan_json(stan_data, data_path)

result_ttest <- pdmp_sample_from_stanmodel(
  model_path_tt, data_path,
  flow = "ZigZag", T = 100000,
  algorithm = "GridThinningStrategy",
  grid_n = 30,
  sticky = TRUE,
  can_stick = c(FALSE, TRUE, FALSE),
  model_prior = bernoulli(0.5),
  parameter_prior = rep(dcauchy(0, 0, rscale), 3)
)
```

```{r}
#| label: ttest-bf
prior_inclusion_prob    <- 0.5
posterior_inclusion_prob <- inclusion_probs(result_ttest)[2]
prior_odds    <- prior_inclusion_prob / (1 - prior_inclusion_prob)
posterior_odds <- posterior_inclusion_prob / (1 - posterior_inclusion_prob)
pdmp_bf <- posterior_odds / prior_odds

analytic_bf <- BayesFactor::extractBF(analytic_results, onlybf = TRUE)

tibble(
  method = c("PDMP", "Analytic"),
  bayes_factor = c(pdmp_bf, analytic_bf)
)
```
